Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-06-04 performance_comparison.tex
[mg] subsec 5.2

We assign a baseline initial covariance matrix $\bP_0 = \bar{\bP}_0$ by selecting standard deviations $\bar{\sigma}_i$ for each state variable. These standard deviations have been chosen based on the typical range of variation of each state.
Assuming the state vector is ordered as described in Section~\ref{sec:vehicle_model}, the values used are $\bar{\boldsymbol{\sigma}}~=~\left[0.1\,\textrm{m/s}, 0.01\,\textrm{m/s}, 0.01\,\textrm{rad/s}, 1\,\textrm{m}, 1\,\textrm{m}, 0.0175\,\textrm{rad}\right]^T$.

In addition to the baseline value $\bar{\bP}_0$, two alternative values are considered: $\bP_0 = 4\bar{\bP}_0$ and $\bP_0 = \frac{1}{4}\bar{\bP}_0$. These correspond, respectively, to doubling and halving the initial standard deviations $\sigma_i$ associated with the state vector. 
A comparison between the trajectories obtained using different values of $\bP_0$ is shown in the right panel of Figure~\ref{fig:ol_sensitivities}.
The black line corresponds to the baseline case with $\bP_0 = \bar{\bP}_0$, while the red and the blue lines represent the cases with $\bP_0 = \frac{1}{4}\bar{\bP}_0$ and $\bP_0 = 4\bar{\bP}_0$, respectively. As reasonably expected, using a larger initial covariance matrix results in a greater covariance matrix after $H$ steps, which in turn leads to a higher back-off term. This is reflected in the blue line, which travels closer to the centerline w.r.t. the other two lines.

\subsection{Robustified constraints comparison}
It is of interest to analyze how a robustified constraint---or a combination of the two described in Sections~\ref{sec:FLC} and~\ref{sec:TLC}---influences the driving style. This analysis compares the not-robust trajectory, obtained without robustified constraints, with those resulting from a robustified track limit constraint (denoted as TLC), a robustified friction limit constraint (denoted as FLC), and a scenario where both constraints are robustified (denoted as TLC+FLC). All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of meeting both constraints.

@@

Figure~\ref{fig:ol_telemetries} presents the variation of four signals with respect to the nominal trajectory, for the three previously mentioned cases. The panels display variations on the longitudinal speed $\De u$ (first panel), wheel steering angle $\De \de$ (second panel), total longitudinal force $\De F_x$ (third panel), and lateral deviation of the center of mass from the track centerline $\De e$ (fourth panel). Except for the $\De u$ panel, each plot includes dashed black lines that indicate the sign of the reference signal-whether it is positive, negative, or numerically zero. These sign indicators are represented as three distinct levels---high, zero, and low---corresponding to positive, zero, and negative values of the reference signal, respectively. The levels are scaled appropriately in each plot to ensure clarity of visualization.

From the first panel, it can be observed that all configurations maintain a lower mean value of the longitudinal speed compared to the reference trajectory, resulting in an increased sector time. Specifically, the configuration with the robustified TLC increases the sector time by 1.66\%, while the configuration with the robustified FLC and with both TLC and FLC robustified increase the sector time by 0.79\% and 2.57\%, respectively.
The two configurations incorporating the robustified track limit constraint (blue and green lines) exhibit, as expected, a significantly altered CoM trajectory. This behavior explains the higher longitudinal speed observed around $\al = 0.71$, which is followed by a sharp reduction.

The second and fourth panels clearly show that the configurations with the robustified track limit constraint tend to follow a path closer to the centerline. In particular, for most of the sector, the variation in lateral displacement $\De e$ and the lateral displacement $e$ itself exhibit opposite signs, indicating a corrective behavior of the steering angle $\de$ that pulls the vehicle toward the centerline.

The third panel, which shows the variation of the total longitudinal force $\De F_x$, is consistent with the $\De u$ panel. In particular, the configurations with the robustified TLC exhibit a higher longitudinal force while approaching Turn 1, in the curvilinear abscissa interval $\al\in\left[0.707, 0.712\right]$, which results in a higher longitudinal speed. Subsequently, these two configurations apply a lower longitudinal force during braking compared to the nominal case, indicating more intense braking. This behavior explains the sharp reduction in longitudinal speed observed thereafter.

It can be interesting to notice that, in the third panel, only the configuration with both the constraints robustified (blue line) needs to reduce the longitudinal force during the high speed turn, in the curvilinear abscissa interval $\al\in\left[0.75, 0.76\right]$, to meet both the constraints. 
The vehicle constrained by the track limit condition is forced to follow a wider trajectory, which entails greater lateral acceleration and, consequently, a higher lateral force demand---resulting in increased tire utilization. Due to the limitations imposed by the robust friction limit constraint, it is necessary to reduce the available accelerating force.

After analyzing the variations of four key signals with respect to the nominal solution, we now turn our attention to the evaluation of axle saturation, which provides insight into how close each configuration operates to the tire grip limits.
Figure~\ref{fig:ol_saturation} presents the quantity $\left[ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2\right]\frac{1}{Z_j^2(\bx,\bu)}$, where $j=1$ refers to the front axle (first panel) and $j=2$ the rear axle (second panel), plotted for the three previously introduced configurations (color-coded as before) and for the nominal solution (black dashed line). This quantity indicates how close each configuration operates to the friction limit; it equals 1 when the point $\left(X_j, Y_j\right)$ lies exactly on the friction ellipse defined by $Z_j$.

@@

As expected, the two configurations with the robustified friction limit constraint (red and blue lines) never reach a value of 1, due to the presence of the back-off term, which enforces a safety margin from the friction limit. 
From this plot, it can be observed that the configurations with the robustified track limit constraint (green and blue lines) follow significantly different trajectories compared to the nominal solution, resulting in distinct ground force demands. This behavior is noticeable in turn 1 and even more clearly in turn 2. Specifically, these two configurations begin to experience high axle loads approximately 12\,m before the nominal trajectory and sustain them for an additional 4.6\,m (TLC, green line) and 9.2\,m (TLC+FLC, blue line) beyond the nominal reference.

>>>>

We assign a baseline initial covariance matrix $\bP_0 = \bar{\bP}_0$ by selecting standard deviations $\bar{\sigma}_i$ for each state variable. These standard deviations have been chosen based on the typical range of variation of each state.
Assuming the state vector is ordered as described in Section~\ref{sec:vehicle_model}, the values used are $\bar{\boldsymbol{\sigma}}~=~\left[0.1\,\textrm{m/s}, 0.01\,\textrm{m/s}, 0.01\,\textrm{rad/s}, 1\,\textrm{m}, 1\,\textrm{m}, 0.0175\,\textrm{rad}\right]^T$.
In addition to the baseline value $\bar{\bP}_0$, two alternative values are considered: $\bP_0 = 4\bar{\bP}_0$ and $\bP_0 = \frac{1}{4}\bar{\bP}_0$. These correspond, respectively, to doubling and halving the initial standard deviations $\sigma_i$ associated with the state vector.

A comparison between the trajectories obtained using different values of $\bP_0$ is shown in the right panel of Figure~\ref{fig:ol_sensitivities}.
The black line corresponds to the baseline case with $\bP_0 = \bar{\bP}_0$, while the red and the blue lines represent the cases with $\bP_0 = \frac{1}{4}\bar{\bP}_0$ and $\bP_0 = 4\bar{\bP}_0$, respectively. As reasonably expected, using a larger initial covariance matrix results in a greater covariance matrix after $H$ steps, which in turn leads to a higher back-off term. This is reflected in the blue line, which travels closer to the centerline with respect to the other two lines.

\subsection{Robustified constraints comparison}
%It is of interest to analyze how a robustified constraint---or a combination of the two described in Sections~\ref{sec:FLC} and~\ref{sec:TLC}---influences the driving style. 
We now investigate how the inclusion of different kinds of robustified constraints---individually or in combination, as described in Sections~\ref{sec:FLC} and~\ref{sec:TLC}---influences the driving style. 
We compare the nominal feed-forward optimal trajectory, obtained without robustified constraints, with those resulting from a robustified track limit constraint (denoted as TLC), a robustified friction limit constraint (denoted as FLC), and a scenario where both constraints are robustified (denoted as TLC+FLC). All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of satisfying both constraints.

@@

%---Delta key signals

The comparison between the nominal case and the three robustified cases is illustrated in Figure~\ref{fig:ol_telemetries}, in terms of the deviation of selected signals from to their nominal counterparts.
The panels display variations on the longitudinal speed $\De u$ (first panel), wheel steering angle $\De \de$ (second panel), total longitudinal force $\De F_x$ (third panel), and lateral deviation of the center of mass (CoM) from the track centerline $\De e$ (fourth panel). 
Except for the $\De u$ panel, each plot includes dashed black lines that indicate the sign of the nominal signal. 
These indicators are shown at three distinct vertical levels---upper, central, and lower---corresponding respectively to positive, near-zero, and negative values of the nominal signal.

%--------1 Considerazioni generali sul settore

From the first panel, it can be observed that all robustified configurations, on average, show a lower longitudinal speed compared to the nominal trajectory, resulting in an increased sector time. Specifically, the configuration with the robustified TLC increases the sector time by 1.66\%, while that with the robustified FLC leads to a 0.79\% increase. When both TLC and FLC are robustified, the sector time increases by 2.57\%.
The two configurations incorporating the robustified TLC (blue and green lines) exhibit, as expected, a significantly altered CoM trajectory.
Ideed, the second and fourth panels clearly show that the TLC and TLC+FLC configurations tend to follow a path closer to the centerline. In particular, for most of the sector, the variation in lateral displacement $\De e$ and the lateral displacement $e$ itself exhibit opposite signs, indicating a corrective behavior of the steering angle $\de$ that pulls the vehicle toward the centerline.

%--------2 Sharp turn

This shift in trajectory and longitudinal speed becomes especially evident when approaching the sharp turn at $\al \in [0.707, 0.712]$. A noticeable increase in longitudinal speed is observed around $\alpha = 0.71$ for the TLC configuration  in the $\De u$ panel. This peak is followed by a sharp drop in speed, indicating intense braking.
The TLC+FLC configuration exhibits a qualitatively similar trend in $\Delta u$, although the peak is significantly lower and the subsequent speed reduction is clearly stronger. This behavior is consistent with the variation of the total longitudinal force $\Delta F_x$ shown in the third panel.
In particular, the TLC and TLC+FLC configurations exhibit a higher longitudinal force while approaching the sharp turn, in the curvilinear abscissa interval $\al\in\left[0.707, 0.712\right]$, resulting in a higher longitudinal speed.
Subsequently, these two configurations apply a lower longitudinal force during braking compared to the nominal case, indicating more intense braking. This behavior explains the sharp reduction in longitudinal speed observed thereafter.

%--------3 High speed turn

Later in the sector, a further deviation in driving style appears during the high-speed turn, specifically around $\alpha \in [0.75, 0.76]$.
As shown in the $\Delta F_x$ panel, the combined configuration TLC+FLC needs to reduce the longitudinal force within this section to meet both the constraints. 
This adaptation is necessary because the TLC forces the vehicle to follow a wider trajectory, which entails greater lateral acceleration and, consequently, higher lateral force demand. To satisfy the FLC under the resulting increase in tire utilization, the available accelerating force must be reduced.
Based on the nominal force sign indicator, this reduction is achieved by partially lifting the throttle.

%---Axle saturation
After analyzing the variations of these four key signals with respect to the nominal solution, we now turn our attention to the evaluation of axle saturation, which provides insight into how close each configuration operates to the tire grip limits. 
In accordance with the FLC formulation introduced in~\ref{eq:adherence_constraint}, we define the \emph{axle saturation ratio} $S_j$ as
\begin{equation}
	S_j = \frac{ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2}{Z_j^2(\bx,\bu)}.
\end{equation}
This quantity indicates how close each configuration operates to the friction limit; $S_j = 1$ when the point $\left(X_j, Y_j\right)$ lies exactly on the friction ellipse, i.e., under full saturation.
Figure~\ref{fig:ol_saturation} shows the values of $S_j$ for the front axle ($j=1$, first panel) and the rear axle ($j=2$, rear panel) for the previously introduced configurations: the nominal case (black dashed line) and the three robustified cases (color-coded as before).

@@

First, we observe that the FLC and TLC+FLC configurations (red and blue lines) never reach full saturation, i.e., $S_j=1$, due to the presence of the back-off term, which enforces a safety margin from the friction limit. 
Second, we point out how the different trajectories followed by the TLC and TLC+FLC configurations (green and blue lines), compared to the nominal and FLC cases, entail distinct ground force demands, which are clearly visible here in terms of the $S_j$ ratio.
This behavior emerges around the sharp turn and becomes even more evident in the high-speed turn. 
Specifically, in the interval $\alpha \in [0.745, 0.762]$, the TLC and TLC+FLC configurations begin to experience high axle loads approximately 12\,m before the nominal trajectory and sustain them for an additional 4.6\,m (TLC, green line) and 9.2\,m (TLC+FLC, blue line) beyond the nominal reference.

%As already observed, the configurations with the robustified track limit constraint (green and blue lines) follow significantly different paths compared to the nominal solution. This deviation results in distinct ground force demands, clearly visible in the saturation index.
%From this plot, it can be observed that the configurations with the robustified track limit constraint (green and blue lines) follow significantly different trajectories compared to the nominal solution, resulting in distinct ground force demands. This behavior is noticeable in turn 1 and even more clearly in turn 2. Specifically, these two configurations begin to experience high axle loads approximately 12\,m before the nominal trajectory and sustain them for an additional 4.6\,m (TLC, green line) and 9.2\,m (TLC+FLC, blue line) beyond the nominal reference.

\subsection{Comparison between open-loop and closed-loop approaches}
